The ChEMBL database as Linked Open Data

### 2012-03-11 article.tex
Small edits to existing form of document, inserted comments about license. Will send seperate email regarding text for ChemSpider to insert into article.

\begin{abstract}
This paper describes the continued conversion of ChEMBL data into RDF triples, using a few existing
ontologis, and demonstrates a few new use cases, taking advantage of its linked nature.
\end{abstract}

@@

Despite the data deluge, finding new, unique, but significant patterns explaining biological
phenomena not yet understood is limited by the availability of data. Such data analysis is
beyond the scope is single data sets; it requires data integration from many sources, for
example, systems biology integrating micro-array differential expression to biological
pathways~\cite{}. Another prominent example is drug discovery where a new unique chemical entity is
searched with just the right biological properties. This too requires linking many
data sets~\cite{Samwald2011,OpenPHACTS}.

ChEMBL contains biological activities for over a million chemical entities, extracted from
literature, providing a unique resource for drug discovers~\cite{Gaulton2012,Warr2009}. Originally, this
data was available for download and via a web interface. The former requires setting up
the data in a relational database, while the latter limits the machine access to the data.
Two independent teams created RDF triples for the ChEMBL data, resulting in exposure
via Chem2Bio2RDF~\cite{Chen2010} and ChEMBL-RDF~\cite{Willighagen2011}.

@@

The ChEMBL version used in this paper is ChEMBL 13, which was released on 29 February 2012.
RDF triples were created from a local MySQL database to which the SQL data dump of ChEMBL was
inserted, using a PHP script, available from source code hosting service
GitHub~\citep{ChEMBLRDFGitHub}, a process outlined earlier in a best practices note by
the international Health Care and Life Sciences XXX Working Group~\cite{Marshall2012}.

During the life time of this project, the ontologies used in ChEMBL-RDF have changed,
initially using a custom, ad-hoc ontology, but increasingly with community-proposed
ontologies, making the RDF more interoperable. The ChEMBL-RDF 13 uses ontology
standards like Bibliography Ontology for Citation Typinh Ontology for liteature
references, and domain ontologies like the Protein Ontology and the Chemical Information
Ontology. Throughout this paper, various prefixes are used to simplify RDF output, outlined
in Table~\ref{namespaces}.

@@

To expose that ChEMBL-RDF data two approaches have been adopted. First, a SPARQL end point
hosted at Uppsala University, using the Open Source Virtuoso software. Use is free, but the
querying is capped, based on the estimated computational effort. Second, resources have
been made dereferencible using the Kasabi platform. This platform also provides a SPARQL
end point, but requires a user account.

@@

Antony: if you can contribute ChemSpider IDs for the ChEMBL compounds ...

>>>>

\begin{abstract}
This paper describes the continued conversion of ChEMBL data into RDF triples, using a few existing
ontologies, and demonstrates a few new use cases, taking advantage of its linked nature.
\end{abstract}

@@

Despite the data deluge, finding new, unique, but significant patterns explaining biological
phenomena not yet understood is limited by the availability of data. Such data analysis is
beyond the scope of single data sets; it requires data integration from many sources, for
example, systems biology integrating micro-array differential expression to biological
pathways~\cite{}. Another prominent example is drug discovery where a new unique chemical entity is
searched with just the right biological properties. This too requires linking many
data sets~\cite{Samwald2011,OpenPHACTS}.

AJW> CIte here the Open PHACTS paper submitted next week: Open PHACTS: Semantic interoperability for drug discovery
Antony J. Williams, Lee Harland, Paul Groth, Stephen Pettifer, Christine Chichester, Egon L. Willighagen, Chris T. Evelo, Niklas Blomberg, Gerhard Ecker, Carole Goble and Barend Mons, Drug Disc. Today. Submitted for publication. Hopefully reviewed and with DOI available by the time this article is accepted.

ChEMBL contains biological activities for over a million chemical entities, extracted from
literature, providing a unique resource for drug discovers~\cite{Gaulton2012,Warr2009}. It is updated on a fairly frequent basis as the existing data is further curated and new data is added. Originally, this
data was available for download and via a web interface. The former requires setting up
the data in a relational database, while the latter limits the machine access to the data.
Two independent teams created RDF triples for the ChEMBL data, resulting in exposure
via Chem2Bio2RDF~\cite{Chen2010} and ChEMBL-RDF~\cite{Willighagen2011}.

@@

The ChEMBL version used in this paper is ChEMBL 13, which was released on 29 February 2012.
RDF triples were created from a local MySQL database to which the SQL data dump of ChEMBL was
inserted, using a PHP script, available from the source code hosting service
GitHub~\citep{ChEMBLRDFGitHub}, a process outlined earlier in a best practices note by
the international Health Care and Life Sciences XXX Working Group~\cite{Marshall2012}.

During the life time of this project, the ontologies used in ChEMBL-RDF have changed,
initially using a custom, ad-hoc ontology, but increasingly with community-proposed
ontologies, making the RDF more interoperable. The ChEMBL-RDF 13 dataset uses ontology
standards such as the Bibliography Ontology for Citation Typing Ontology for literature
references, and domain ontologies like the Protein Ontology and the Chemical Information
Ontology. Throughout this paper, various prefixes are used to simplify the RDF output and they are outlined
in Table~\ref{namespaces}.

@@

To expose that ChEMBL-RDF data two approaches have been adopted. First, a SPARQL end point
hosted at Uppsala University, using the Open Source Virtuoso software. Use is free, but the
querying is capped, based on the estimated computational effort. Second, resources have
been made dereferencable using the Kasabi platform. This platform also provides a SPARQL
end point, but requires a user account.

@@

Antony: if you can contribute ChemSpider IDs for the ChEMBL compounds ... 
AJW> CC-BY-SA 3.0 licensing approved for the mappings between ChEMBL and ChemSpider. Please use the Open PHACTS data provided and attribution to ChEMBL, OPS and ChemSPider.
